Employee Attrition: What makes an employee leave?

### 2017-10-19 LatexPaper/paper.tex
updated section 3, attrition as seen in civil service workers data background

\paragraph{The U.S. Office of Personnel Management (OPM) serves as the central Human Resources department for all Federal agencies, including the management of federal agency health insurance and retirement benefits. Their oversight of policy implementation as well as being a general resource for all agency Human Resource departments makes their employment data of particular interest for this paper. OPM regularly captures a wide range of data on the millions of federal civil service workers, and we used several of their data sources that focus on attrition. These data [7], include variables such as job level, locality, salary, length of service, and basic reason for termination of employment. Important for understanding our findings in the correct context are the following points:}

\subparagraph{October 2014 – September 2015: The date range used to filter the final dataset (one full calendar year).}

\subparagraph{General Schedule Pay Scale (GS): This is the centralized pay scale used by many civil service agencies which covers the majority of white collar professionals. Whether or not a person is paid under the GS, the OPM converts the level he or she is paid under to the GS for data collection purposes. The scale numbers 1-15 and there are ten “steps” within each of those levels.}

\subparagraph{Locality adjustments: To remain competitive with industry salaries, the GS operates a locality adjustment scale that adds a particular percentage to a person’s salary, based on city of occupation alone.}

\subparagraph{Federal Employees Retirement System (FERS): Beginning on January 1, 1987, new civil employees are paid under FERS. This mix of Social Security, a Basic Benefit Plan, and a Thrift Savings Plan helps make civil service positions stand out from many current private sector jobs. We account for this factor in our research, taking into consideration the confounding effect this has on length of service (LOS). At the time of this paper, the pension is based on salary and length of service.}

\subparagraph{Reasons for separation: OPM uses twelve different categorical values to capture separation data.  A full list can be found with the original dataset [22], but it is worth noting for the reader that the events recorded include disability, death, various types of transfers, and several types of retirement.}

@@

\paragraph{Within the original OPM data, six new attributes were created through aggregation or calculation amongst various attributes: 1) SEP Count by Date \& Occupation – Total number of separations (of any type) for a given Date and Occupation; 2) SEP Count by Date \& Location – Total number of separations (of any type) for a given Date and Location; 3) Industry Average Salary – Average salary amongst non-separated employees, grouped by quarter, occupation, pay grade, and work schedule; 4) Lower Limit Age – Youngest age within each age level category; 5) Years to Retirement – Based on FERS retirement eligibility baseline of 57 years of age [8]; and 6) Salary Over/Under Industry Average – Difference between computed average salary of non-separated employees and actual salary for each observation. Note also another 1,293 observations were removed after calculating industry average salary as they had no matching non-separation observations (matched on quarter, occupation type, pay plan /grade, and work schedule), which were utilized to ensure realistic salary averages.}

>>>>

\paragraph{The U.S. Office of Personnel Management (OPM) serves as the central Human Resources department for all Federal agencies, including the management of federal agency health insurance and retirement benefits. Their oversight of policy implementation as well as being a general resource for all agency Human Resource departments makes their employment data of particular interest for this paper. OPM releases on a quarterly basis updated separation data on its over two million federal civil service workers.  We used these anonymized data sets to create a full year of separation data, spanning October 2014 - September 2015.   These data [7], include the following variables: 1) age group, captured categorically in 5 year increments (AGELVL and AGELVLT), 2) agency type and subtype information (AGYTYP, AGYTYPT, AGY, AGYT, AGYSUB, and AGYSUBT), 3) gender (GENDER and GENDERT), 4) General Schedule Grade, or equivalent, dictating salary level (GSEGRD, 5) geographical location of the employee down to the state level (LOCTYP, LOCTYPT, LOC, LOCT), 5)length of service as a government employee captured in one year increments until five years was achieved, then every five after that (LOSLVL and LOSLVLT), 6)occupation type  in regards to white or blue collar, occupation family, and in some cases, down to the particular occupation itself (i.e. "chaplain") (OCCTYP, OCCTYPT, OCCFAM, OCCFAMT, OCC, OCCT), 7) occupation category among seven main categories (i.e. "Technical" and "Clerical")(PATCO and PATCOT), 8) pay plan and grade for both GSE and non-GSE occupations (PPTYP, PPTYPT, PPGROUP, PPGROUPT, PAYPLAN, PAYPLANT, and PPGRD), 9) salary range, captured categorically starting with < \$20,000 and followed in \$10,000 increments (SALLVL and SALLVLT), 10) reason for separation (SEP and SEPT), and 11) type of appointment and worschedules, denoting factors such as permanence, seasonality, and executive status (TOATYP, TOATYPT, TOA, TOAT, WSTYP, WSTYPT, WORKSCH, and WORKSCT).   }

\subparagraph{There are certain points worth noting to understand our findings in the correct context.  The General Schedule Pay Scale is the centralized pay scale used by many civil service agencies. Whether or not a person is paid under the GS, the OPM converts the level he or she is paid under to the GS for data collection purposes. The scale numbers 1-15 and there are ten “steps” within each of those levels.  Also, to remain competitive with industry salaries, the GS operates a locality adjustment scale that adds a particular percentage to a person’s salary, based on city of occupation alone.}

\subparagraph{ Also, beginning on January 1, 1987,  new civil service employees were paid under the Federal Employees Retirement System (FERS). This mix of Social Security, a Basic Benefit Plan, and a Thrift Savings Plan helps make civil service positions stand out from many current private sector jobs in terms of retention. We account for this factor in our research, taking into consideration the confounding effect this has on length of service (LOS). At the time of this paper, the pension is based on salary and length of service.}

@@

\paragraph{Six new attributes were created through aggregation or calculation: 1) SEP Count by Date \& Occupation – Total number of separations (of any type) for a given Date and Occupation; 2) SEP Count by Date \& Location – Total number of separations (of any type) for a given Date and Location; 3) Industry Average Salary – Average salary amongst non-separated employees, grouped by quarter, occupation, pay grade, and work schedule; 4) Lower Limit Age – Youngest age within each age level category; 5) Years to Retirement – Based on FERS retirement eligibility baseline of 57 years of age [8]; and 6) Salary Over/Under Industry Average – Difference between computed average salary of non-separated employees and actual salary for each observation. Note also another 1,293 observations were removed after calculating industry average salary as they had no matching non-separation observations (matched on quarter, occupation type, pay plan /grade, and work schedule), which were utilized to ensure realistic salary averages.}
